---
title: "GEO"
format: html
editor: visual
echo: false
eval: false
---

# GEO - Gene Expression Omnibus (GEO)

Data repository from NCBI with functional genomic data.

-   Platform: an array or sequencer.
-   Sample: sample source, protocols used, and expression data.
-   Series: related samples.
-   DataSet: curated collection of biologically and statistically comparable GEO Samples.
-   Profile: expression measurements for an individual gene across all Samples in a DataSet.

## 1. Datasets structure

-   assayData: raw or processed intensities (samples in columns).
-   phenoData: metadata (samples in rows).
-   featureData: annotation.
-   annotation: platform name.

![\[1\]: Figure from Greg Tucker-Kellogg](images/ExpressionSet.png)

### GEO Files

-   SOFT: information about the experiment (MIAME), estimates for each gene, transcript.
-   MINiML: XML format.
-   Series matrix file: directly to a spreadsheet.
-   \`.CEL\` (Supplemental files): Original, unprocessed data.
-   GSE object: sample(GSM, GSMList) + platform(GPL, GPLList)
-   \`GDS2eSet\` ExpressionSet with processed data.

## 1. Loading packages

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("GEOquery")
BiocManager::install("limma")
BiocManager::install("pheatmap")
BiocManager::install("org.Hs.eg.db")
pacman::p_load(GEOquery, tidyverse, ggrepel, survminer, limma, pheatmap, org.Hs.eg.db)

# tips: browseVignettes('packagename')
# tips: methods(class=class(foo))
```

```{r}
# Fastest approach assuming the counts doesnt contain errors 

htlv_id = "GSE34870"
htlv_tempdir="/Users/denriquez/Documents/tempdev"
gse = getGEO(
  htlv_id, 
  GSEMatrix=TRUE, 
  destdir=htlv_tempdir
  )

# To know about the processing or normalization (Affymetrix or Agilent)
gse[[1]]$data_processing
gse[[1]]$platform_id

# Exploring
class(gse)
methods(class=class(gse))
length(gse)

# Subsetting to 1
GSE34870<-gse[[1]]

# Exploring
class(GSE34870)
methods(class=class(GSE34870))

# Exploring more
pData(GSE34870)[1:46,] # Data information or Phenotype data

fData(GSE34870) # Gene annotation

exprs(GSE34870)[1,] # Expression data

varLabels(GSE34870) # Metadata

GSE34870$supplementary_file #Raw files

annotation(GSE34870) # Chip information

experimentData(GSE34870) #Experimental info

featureData(GSE34870) #Probeset data

#To check if raw data matchs with metadata
pd <-pData(GSE34870)
pd['supp_name'] <- str_split(pd$supplementary_file,"/") |>
  map_chr(tail,1)
pd['supp_name'] 

```

## 3. Normalisation

### Affymetrix

RMA: Robust Multiarray Average (Affymetrix arrays), this includes log2 at the last step.

AffyBatch -\> ExpressionSet

-   Background correction
-   Normalisation
-   Summarisation

```{r}
pData(gse)$data_processing[1] 

summary(exprs(gse)) # rarely more than 16
# exprs(gse) <- log2(exprs(gse))

boxplot(exprs(gse),outline=F)
```

## 4. Linear models

Generate a model to specify the design and fit the model to the design.

`y ~ x + 0`: The "0" will remove the intercept.

```{r}
library(limma)

design_htlv <- model.matrix( ~ GSE34870[['cell line:ch1']])
colnames(design_htlv)[2] <- "model_htlv"

# To check our model
design_htlv

# Fit a linear model 
fit_htlv <-lmFit(GSE34870, design_htlv)

# Empirical Bayes correction in limma: distribution across genes to calculate a robust test statistic
fitted.ebayes <- eBayes(fit_htlv)

# Extracting DE genes
topTable(fitted.ebayes)

# Decide test - which features pass the test criteria (Benjamini-Hochberg p=0.05, no log fold change cutoff)

summary(decideTests(fitted.ebayes[, "model_htlv"], lcf=1))
```

## 5. Data wrangling

```{r}
sample_info = pData(gse)
head(sample_info, n=20)

sample_wt = sample_info |>
  select(characteristics_ch1, characteristics_ch1.2) |>
  filter(characteristics_ch1.2=="treatment: untreated for 6 hours")
```

```{r}
library(stringr)

sample_wt$group <- ""
for(i in 1:nrow(sample_wt)){
  if(str_detect(sample_wt$characteristics_ch1[i], "MT-2")){sample_wt$group[i] <- "MT-2"}
  if(str_detect(sample_wt$characteristics_ch1[i], "MT-4")){sample_wt$group[i] <- "MT-4"}
  
}

```

## 6. Sample clustering and exploratory PCA

```{r}
library(pheatmap)
## argument use="c" stops an error if there are any missing data points

corMatrix <- cor(exprs(gse),use="c")
pheatmap(corMatrix) 

colnames(corMatrix)
rownames(sample_info)
# Supplementary files (AGILENT)
filePaths <- getGEOSuppFiles("GSE34870") # this may fail

gse <- gse[[1]]
for (i in 1:length(gse$supplementary_file)) {
  url <- gse$supplementary_file[i]
  destfile <- file.path(htlv_tempdir, basename(url))
  # Download the file
  #download.file(url, destfile, mode = "wb")
  download.file(url, destfile, mode = "wb", method = "curl")
  #system(paste("wget -O", destfile, url))
  # Extract the tar file if it's a tar archive
  #untar(destfile, exdir = htlv_tempdir)
}










pd <- pData(gse)
pd['cel_file'] <- str_split(pd$supplementary_file,"/") |>
  map_chr(tail,1)


pacman::p_load(oligo)
pacman::p_load(oligoClasses)

GSE34870_celdata <- read.celfiles(list.celfiles('htlv_id',full.names=TRUE,listGzipped=TRUE))

```

```{r}
htlv_id = "GSE34870"
gse = getGEO(htlv_id, GSEMatrix=FALSE) # parse the SOFT (less errors and more info, also locally filename="*.soft")

# Exploring
class(gse)
methods(class=class(gse))
names(GSMList(gse))
names(Meta(gse))
names(GPLList(gse))








### CEL files for raw data (FeatureSet)
filePaths <- getGEOSuppFiles('GSE34870')
library(oligo)
library(oligoClasses)
gse33146_celdata <- read.celfiles(list.celfiles('GSE33146',full.names=TRUE,listGzipped=TRUE))
image(gse33146_celdata[,1])

```

## Further Info

-   (Tutorial by Mark Dunning)\[https://sbc.shef.ac.uk/geo_tutorial/tutorial.nb.html\]

-   (Tutorial by Lindseynicer)\[https://github.com/Lindseynicer/How-to-analyze-GEO-microarray-data\]

-   (Repository in Spanish)\[https://github.com/u-genoma/BioinfinvRepro\]

-   (Deeply explained)\[https://gtk-teaching.github.io/Microarrays-R/\]