---
title: "GEO"
format: html
editor: visual
echo: false
eval: false
---

# GEO - Gene Expression Omnibus (GEO)

Data repository from NCBI with functional genomic data.

## 1. Definitions

-   Platform: an array or sequencer.
-   Sample: sample source, protocols used, and expression data.
-   Series: related samples.
-   DataSet: curated collection of biologically and statistically comparable GEO Samples.
-   Profile: expression measurements for an individual gene across all Samples in a DataSet.

## 2. Datasets structure

-   assayData: raw or processed intensities (samples in columns).
-   phenoData: metadata (samples in rows).
-   featureData: annotation.
-   annotation: platform name.

![\[1\]: Figure from Greg Tucker-Kellogg](images/ExpressionSet.png)

## 3. GEO Files

-   SOFT: information about the experiment (MIAME), estimates for each gene, transcript.
-   MINiML: XML format.
-   Series matrix file: directly to a spreadsheet.
-   `.CEL` (Supplemental files): Original, unprocessed data.
-   GSE object: sample(GSM, GSMList) + platform(GPL, GPLList)
-   `GDS2eSet` ExpressionSet with processed data.

![\[2\]: GSE55851](images/paste-1.png)

## 4. Normalisation

### Affymetrix

RMA: Robust Multiarray Average (Affymetrix arrays), this includes log2 at the last step.

AffyBatch -\> ExpressionSet

-   Background correction: removal of background noise (probe pairs, a perfect match (PM), andsingle base mismatch (MM): Affymetrix’s own MAS5 algorithm, RMA ALGORITHM, gcrma algorithm)
-   Normalisation:  quantile normalisation, data on a common empirical distribution
-   Summarisation

## Tutorials

### 1. Loading packages
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("GEOquery")
BiocManager::install("limma")
pacman::p_load(GEOquery, tidyverse, ggrepel, survminer, limma, oligo)
pacman::p_load(pheatmap, tidyverse, ggrepel, tidyplots)
# tips: browseVignettes('packagename')
# tips: methods(class=class(foo))
# for vscode: pip3 install -U radian
```

### 2. Scenario A: Assuming everything is fine (data normalised)
```{r}
# GEO ID and folder
htlv_id = "GSE55851"
htlv_tempdir=".temp/"

# 1. obtaining processed data
gse2 = getGEO(
  htlv_id, 
  GSEMatrix=TRUE, 
  destdir=htlv_tempdir)

# 2. exploring: number of platforms
length(gse2)

# 3. subsetting
gse2 <- gse2[[1]]

# 4. further exploration
pData(gse2) # Sampledata
fData(gse2) # gene annotation
exprs(gse2) # expression data

head(pData(gse2))
colnames(pData(gse2))
gse2@annotation

# 5. metadata - extract relevant data
pinfo = pData(gse2) |>
  dplyr::rename(typex="characteristics_ch1.1", age="characteristics_ch1.2", gender="characteristics_ch1.3") |>
  dplyr::select(typex, age, gender) |>
  dplyr::mutate(across(everything(), ~ sub(".*: ", "", .)))

pinfo$ID <- rownames(pinfo)
pinfo

# 6. Check if normalised
oligo::hist(log2(exprs(gse2)))

boxplot(log2(exprs(gse2)), outline=F) #not normalised

#Unsupervised clustering (understand the sources of variation and outliers)
rownames(pinfo) <- colnames(cor(exprs(gse2), use="c"))

#Check IDs
table(rownames(pinfo), pinfo$ID)

pheatmap(cor(exprs(gse2), use="c"), annotation_col=pinfo[,c(3,1)]) #c stops if any missing data point, "complete.obs"

# PCA - mandatory to transpose
pca <- prcomp(t(exprs(gse2)))

cbind(pinfo, pca$x) |>
tidyplot(x=PC1, y=PC2, col=typex, label=age, shape=gender) |>
add_data_points(rasterize = FALSE, rasterize_dpi = 300) |>
add_data_labels_repel() |>
adjust_size(width = 100, height = 100, unit = "mm")

#remove outliers: gse <- gse[,-outlier_samples]

#---- Normalisation
# Load required libraries
library(limma)

# Step 2: Background correction
gsm2 <- exprs(gse2)

gsm2_background<-limma::backgroundCorrect(gse2, method = "normexp")

# Step 3: Normalize the data
agilent_data_norm <- limma::normalizeWithinArrays(agilent_data_norm, method = "loess")

agilent_data_norm <- limma::normalizeBetweenArrays(gsm2_background, method = "quantile")
agilent_data_norm

boxplot(log2(agilent_data_norm), outline=F)
pca <- prcomp(t(log2(agilent_data_norm)))
library(tidyplots)
cbind(pinfo, pca$x) |>
tidyplot(x=PC1, y=PC2, col=typex, label=age, shape=gender) |>
add_data_points(rasterize = FALSE, rasterize_dpi = 300) |>
add_data_labels_repel() |>
adjust_size(width = 100, height = 100, unit = "mm")

library(pheatmap)
pheatmap(cor(agilent_data_norm, use="c"), annotation_col=pinfo[,c(3,1)])

# Step 4: Log transformation (if needed)
agilent_data_log <- log2(agilent_data_norm$M)

GSE46518_set <- rma(GSE46518_celdata)
hist(GSE46518_set, target="core")

# Step 3: Within-array normalization (Loess)
agilent_data_norm <- normalizeWithinArrays(agilent_data_bg, method = "loess")

# Step 4: Between-array normalization (Quantile)
agilent_data_norm <- normalizeBetweenArrays(agilent_data_norm, method = "quantile")

# Step 5: Log2 transformation
agilent_data_log <- log2(agilent_data_norm$M)

# Step 6: Save normalized data
write.table(agilent_data_log, file = "normalized_data.txt", sep = "\t", quote = FALSE)
```

```{r}
# Option A: Retrieving GSE objects

htlv_id = "GSE55851"
htlv_tempdir=".temp/"
gse = getGEO(
  htlv_id, 
  GSEMatrix=FALSE, #parse SOFT (more information, less errors, slower)
  destdir=htlv_tempdir)
methods(class=class(gse))
Meta(gse)
#methods
length(gse)
class(gse)
methods(class=class(gse) #GPLList + GSMList + Metadata
names(GSMList(gse))
names(GPLList(gse))

# Option B: Retrieving ExpressionSet objects (Geo dataset)
gse2 = getGEO(
  htlv_id, 
  GSEMatrix=TRUE, 
  destdir=htlv_tempdir)

gse2[[1]]$data_processing
gse2[[1]]$platform_id

# Exploring
class(gse2)
methods(class=class(gse2))
length(gse2)

# Subsetting to 1
GSE55851<-gse2[[1]]

# Exploring
class(GSE55851)
methods(class=class(GSE55851))
length(GSE55851)
varLabels(GSE55851)
GSE55851$supplementary_file

pd <- pData(GSE55851)
pd["cel_file"] <- str_split(GSE55851$supplementary_file, "/") |> map_chr(tail,1)
pd["cel_file"]

# Option C: raw files mostly for affymetryx, ExpressionFeatureSet or genefeatureset
getGEOSuppFiles('GSE55851',makeDirectory = TRUE, baseDir = "./.temp")
library(oligo)
library(oligoClasses)

# Check the platform (GPL) number
Meta(gse)$platform_id

# Agilent: single vs double channel (green and red)
library(limma)
# Step 1: Load Agilent data
files <- list.files("path/to/Agilent/files", pattern = "*.txt", full.names = TRUE)
agilent_data <- read.maimages(files, source = "agilent", green.only = TRUE)

#or

files <- list.files("GSE55851/GSE55851_RAW/", pattern = "^GSM.*.txt.gz", full.names = TRUE)
files
# Read the Agilent files
agilent_data <- read.maimages(files, source="agilent", columns=list(G = "gMedianSignal"), green.only=TRUE)
agilent_data
agilent_data$targets

#If affymetrix
gse55851_celdata <- read.celfiles(paste0('GSE55851/GSE55851_RAW/',pd$cel_file),phenoData=phenoData(gse2))
pData(gse55851_celdata)[,c("geo_accession", "cell type:ch1", "treatment_ch1")]
varLabels(gse2)

```

```{r}
# Ultimate example
htlv_id = "GSE46518"
# Obtain the processed data
gse_GSE46518 = getGEO(
  htlv_id, 
  GSEMatrix=TRUE, 
  destdir="./.temp")
gse_GSE46518[[1]]$supplementary_file

#Obtain data and attach to supp files
pd <- pData(gse_GSE46518[[1]])
pd["cel_file"] <- str_split(gse_GSE46518[[1]]$supplementary_file, "/") |> map_chr(tail,1)
pd["cel_file"]

#Download raw data
getGEOSuppFiles('GSE46518',makeDirectory = TRUE, baseDir = ".temp/")
# Another download option
gse46518 = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE46518&format=file"

pacman::p_load(curl)
curl_download(gse46518, destfile = "./.temp/GSE46518_RAW.tar")

# Untar
untar(".temp/GSE46518_RAW.tar", exdir = "./GSE46518")

# Read CEL files
library(Biobase)  # Required for AnnotatedDataFrame
# Convert the data.frame to an AnnotatedDataFrame
gsepd<- AnnotatedDataFrame(pData(gse_GSE46518[[1]]))
GSE46518_celdata <- read.celfiles(paste0('GSE46518/',pd$cel_file),phenoData=gsepd)

varLabels(GSE46518_celdata)
pData(GSE46518_celdata)[,c("geo_accession", "infection:ch1")]

hist(GSE46518_celdata, target="core")

## RMA
GSE46518_set <- rma(GSE46518_celdata)
hist(GSE46518_set, target="core")

## gcrma, when concern about GC content, affymetryx and low levels
pacman::p_load(gcrma)
GSE46518_gcrma <- gcrma(GSE46518_celdata)
```
```{r}
# Exploring ExpressionSet
pData(GSE55851)[1:5,] # Data information or Phenotype data
fData(GSE55851) # Gene annotation
exprs(GSE55851)[1,] # Expression data
varLabels(GSE55851) # Metadata
GSE55851$supplementary_file #Raw files
annotation(GSE55851) # Chip information
experimentData(GSE55851) #Experimental info
featureData(GSE55851) #Probeset data
#To check if raw data matchs with metadata
pdcheck <-pData(GSE55851)
pdcheck['supp_name'] <- str_split(pdcheck$supplementary_file,"/") |>
  map_chr(tail,1)
pdcheck['supp_name'] 
```


```{r}
pData(GSE46518_celdata)$data_processing[1] 
summary(exprs(GSE46518_celdata)) # rarely more than 16
# exprs(gse) <- log2(exprs(gse))
boxplot(exprs(gse),outline=F)
```

## 4. Linear models

Generate a model to specify the design and fit the model to the design.

`y ~ x + 0`: The "0" will remove the intercept.

```{r}
library(limma)

design_htlv <- model.matrix( ~ GSE46518_celdata[["infection:ch1"]])
colnames(design_htlv)[2] <- "model_htlv"

# To check our model
design_htlv

# Fit a linear model 
fit_htlv <-lmFit(GSE46518_set, design_htlv)

# Empirical Bayes correction in limma: distribution across genes to calculate a robust test statistic
fitted.ebayes <- eBayes(fit_htlv)

# Extracting DE genes
topTable(fitted.ebayes)

# Decide test - which features pass the test criteria (Benjamini-Hochberg p=0.05, no log fold change cutoff)

summary(decideTests(fitted.ebayes[, "model_htlv"], lcf=1))

GSE46518_ex_values <- exprs(GSE46518_celdata)
fData(GSE46518_celdata)
# Define probe set IDs for sex-specific genes
sex_genes <- c("220729_s_at", "205000_at", "205038_at", "205040_at")
sex_gene_expression <- GSE46518_ex_values[sex_genes, ]
head(GSE46518_ex_values)

annotation(GSE46518_celdata)
featureNames(GSE46518_celdata)
gse_GSE46518
gpl <- getGEO("GPL5175")
anno_table <- Table(gpl)
head(anno_table)

```

c("XIST: 220729_s_at (Affymetrix Human Genome U133 Plus 2.0 Array)
RPS4Y1: 205000_at
KDM5D: 205038_at
DDX3Y: 205040_at

```{r}
#vector of sex genes
sex = c("XIST", "RPS4Y1", "KDM5D", "DDX3Y")

## table
str(anno_table)
selected_rows <- anno_table[grep("DDX3Y|KDM5D|RPS4Y1|XIST", anno_table$mrna_assignment, ignore.case = TRUE), ]

sex_id = selected_rows$ID

sex_id

###
head(GSE46518_ex_values)
# Install required packages if not already installed
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

if (!require("pd.huex.1.0.st.v2", quietly = TRUE)) 
  BiocManager::install("pd.huex.1.0.st.v2")

# Load the required libraries
library(pd.huex.1.0.st.v2)
library(AnnotationDbi)

# Create the annotation database object
huex_annot <- pd.huex.1.0.st.v2

getdb(huex_annot)
methods(class=class(huex_annot))
str(huex_annot)
# Get probe IDs using the correct method
pkgenv <- as.env(huex_annot)

# Function to retrieve annotations
get_probe_annotation <- function(probe_ids) {
  annotations <- select(huex_annot, 
                       keys = probe_ids,
                       columns = c("PROBEID", "SYMBOL", "GENENAME", "ENTREZID", "ENSEMBL"),
                       keytype = "PROBEID")
  return(annotations)
}

# Example usage:
# Replace with your specific probe IDs of interest
probe_ids <- head(available_keys, 10)
annotations <- get_probe_annotation(probe_ids)
print(annotations)

# To see all available annotation types:
print(available_columns)
```
## 5. Data wrangling

```{r}
sample_info = pData(GSE46518_set)
head(sample_info, n=20)

sample_wt = sample_info |>
  select(characteristics_ch1, characteristics_ch1.2) |>
  filter(characteristics_ch1.2=="treatment: untreated for 6 hours")
```

```{r}
library(stringr)

sample_wt$group <- ""
for(i in 1:nrow(sample_wt)){
  if(str_detect(sample_wt$characteristics_ch1[i], "MT-2")){sample_wt$group[i] <- "MT-2"}
  if(str_detect(sample_wt$characteristics_ch1[i], "MT-4")){sample_wt$group[i] <- "MT-4"}
  
}

```

## 6. Sample clustering and exploratory PCA

```{r}
library(pheatmap)
## argument use="c" stops an error if there are any missing data points

corMatrix <- cor(exprs(gse),use="c")
pheatmap(corMatrix) 

colnames(corMatrix)
rownames(sample_info)
# Supplementary files (AGILENT)
filePaths <- getGEOSuppFiles("GSE34870") # this may fail

gse <- gse[[1]]
for (i in 1:length(gse$supplementary_file)) {
  url <- gse$supplementary_file[i]
  destfile <- file.path(htlv_tempdir, basename(url))
  # Download the file
  #download.file(url, destfile, mode = "wb")
  download.file(url, destfile, mode = "wb", method = "curl")
  #system(paste("wget -O", destfile, url))
  # Extract the tar file if it's a tar archive
  #untar(destfile, exdir = htlv_tempdir)
}










pd <- pData(gse)
pd['cel_file'] <- str_split(pd$supplementary_file,"/") |>
  map_chr(tail,1)


pacman::p_load(oligo)
pacman::p_load(oligoClasses)

GSE34870_celdata <- read.celfiles(list.celfiles('htlv_id',full.names=TRUE,listGzipped=TRUE))

```

```{r}
htlv_id = "GSE34870"
gse = getGEO(htlv_id, GSEMatrix=FALSE) # parse the SOFT (less errors and more info, also locally filename="*.soft")

# Exploring
class(gse)
methods(class=class(gse))
names(GSMList(gse))
names(Meta(gse))
names(GPLList(gse))








### CEL files for raw data (FeatureSet)
filePaths <- getGEOSuppFiles('GSE34870')
library(oligo)
library(oligoClasses)
gse33146_celdata <- read.celfiles(list.celfiles('GSE33146',full.names=TRUE,listGzipped=TRUE))
image(gse33146_celdata[,1])

```

## Further Info

-   (Tutorial by Mark Dunning)\[https://sbc.shef.ac.uk/geo_tutorial/tutorial.nb.html\]

-   (Tutorial by Lindseynicer)\[https://github.com/Lindseynicer/How-to-analyze-GEO-microarray-data\]

-   (Repository in Spanish)\[https://github.com/u-genoma/BioinfinvRepro\]

-   (Deeply explained)\[https://gtk-teaching.github.io/Microarrays-R/\]